---
title: "Team 3 - Hazardous Celestial Objects: Full Analysis"
author: "Alejandro Arroyo, Ryujin Bupp, Jose Placeres, Fernando Ricaurte"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

# Libraries

```{r libraries}
library(dplyr)
library(ggplot2)
library(tidyr)
library(knitr)
library(kableExtra)
library(caret)
library(randomForest)
library(pROC)
library(glmnet)
library(leaps)
library(MASS)
library(rpart)
library(rpart.plot)
library(gbm)
library(scales)
library(tibble)
library(corrplot)
library(GGally)
```

---

# Data Cleaning and Preprocessing

## Loading the Raw Dataset

```{r load-raw-data}
# Dataset source:
# https://www.kaggle.com/datasets/seniruepasinghe/near-earth-comets-classification-dataset
# Download manually and place in your working directory before running

asteroids <- read.csv(
  "nasa_neows_1950_2025.csv",
  header           = TRUE,
  stringsAsFactors = FALSE
)

str(asteroids)
summary(asteroids)
head(asteroids)
```

The original full dataset contains 375,659 rows and 41 columns.

## Removing Missing Values

```{r remove-na}
asteroids_clean <- na.omit(asteroids)
str(asteroids_clean)
```

## Removing Redundant Columns

```{r remove-columns}
# Remove Name and Equinox in favor of Neo Reference ID
asteroids_clean <- asteroids_clean[, !(names(asteroids_clean) %in% c("Name", "Equinox"))]

# Remove superfluous distance/velocity unit columns to maintain
# a single consistent unit of measurement
superfluous <- c(
  "Est.Dia.in.KM.min.", "Est.Dia.in.M.min.", "Est.Dia.in.M.max.",
  "Est.Dia.in.Miles.min.", "Est.Dia.in.Miles.max.",
  "Est.Dia.in.Feet.min.", "Est.Dia.in.Feet.max.",
  "Relative.Velocity.km.per.sec", "Miles.per.hour",
  "Miss.Dist..Astronomical.", "Miss.Dist..lunar.", "Miss.Dist..miles."
)

asteroids_clean <- asteroids_clean[, !(names(asteroids_clean) %in% superfluous)]

str(asteroids_clean)
colnames(asteroids_clean)
```

## Filtering to Earth-Orbiting Objects

```{r filter-earth}
unique(asteroids_clean$Orbiting.Body)
asteroids_clean <- asteroids_clean[asteroids_clean$Orbiting.Body == "Earth", ]
unique(asteroids_clean$Orbiting.Body)
str(asteroids_clean)
```

## Removing Duplicate Objects

```{r remove-duplicates}
sum(duplicated(asteroids_clean$Neo.Reference.ID))
asteroids_clean <- asteroids_clean[!duplicated(asteroids_clean$Neo.Reference.ID), ]
sum(duplicated(asteroids_clean$Neo.Reference.ID))
str(asteroids_clean)
colnames(asteroids_clean)
```

## Final Null Check

```{r null-check}
asteroids_clean <- na.omit(asteroids_clean)
colSums(is.na(asteroids_clean))
```

---

# Exploratory Data Analysis

```{r eda-class-distribution}
ggplot(asteroids_clean, aes(x = as.factor(Hazardous), fill = as.factor(Hazardous))) +
  geom_bar() +
  theme_minimal() +
  labs(title = "Class Distribution: Hazardous vs Non-Hazardous",
       x     = "Hazardous",
       y     = "Count",
       fill  = "Hazardous") +
  scale_fill_manual(values = c("TRUE" = "red", "FALSE" = "steelblue"))
```

```{r eda-miss-dist}
ggplot(asteroids_clean, aes(x = Miss.Dist..kilometers.)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Miss Distance",
       x     = "Miss Distance (km)",
       y     = "Count")
```

```{r eda-velocity-vs-dist}
ggplot(asteroids_clean,
       aes(x = Relative.Velocity.km.per.hr,
           y = Miss.Dist..kilometers.)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", color = "red") +
  theme_minimal() +
  labs(title = "Velocity vs Miss Distance",
       x     = "Relative Velocity (km/hr)",
       y     = "Miss Distance (km)")
```

```{r eda-magnitude-vs-dist}
ggplot(asteroids_clean,
       aes(x = Absolute.Magnitude,
           y = Miss.Dist..kilometers.)) +
  geom_point(alpha = 0.4, color = "steelblue") +
  geom_smooth(method = "lm", color = "red") +
  theme_minimal() +
  labs(title = "Absolute Magnitude vs Miss Distance",
       x     = "Absolute Magnitude",
       y     = "Miss Distance (km)")
```

```{r eda-correlation}
numeric_data <- select_if(asteroids_clean, is.numeric)
cor_matrix   <- cor(numeric_data, method = "spearman", use = "complete.obs")

corrplot(
  cor_matrix,
  method = "color",
  type   = "upper",
  order  = "hclust",
  tl.col = "black",
  tl.cex = 0.7,
  tl.srt = 45
)
```

```{r eda-pairs}
subset_vars <- asteroids_clean %>%
  dplyr::select(Miss.Dist..kilometers.,
                Relative.Velocity.km.per.hr,
                Absolute.Magnitude,
                Orbital.Period,
                Minimum.Orbit.Intersection)

ggpairs(subset_vars)
```

---

# Variable Selection — Logistic Regression, Ridge, and LASSO

## Data Preparation

```{r step2-load}
# Using asteroids_clean already created in the preprocessing section
# Drop all non-predictive columns once for use across all models
shared_drop <- c("Date", "Neo.Reference.ID", "Close.Approach.Date",
                 "Orbit.Determination.Date", "Epoch.Date.Close.Approach",
                 "Orbit.ID", "Perihelion.Time", "Orbiting.Body")

model_data <- asteroids_clean[, !(names(asteroids_clean) %in% shared_drop)]

# Convert all predictors to numeric
predictor_names <- setdiff(names(model_data), "Hazardous")
model_data <- model_data %>%
  mutate(across(all_of(predictor_names), ~ as.numeric(as.character(.))))

# hco_data — numeric Hazardous for logistic regression, Ridge, LASSO
hco_data           <- model_data
hco_data$Hazardous <- ifelse(as.character(model_data$Hazardous) %in% c("True", "TRUE"), 1, 0)

# comets_model — factor Hazardous for RF, Boosting, Bagging
comets_model           <- model_data
comets_model$Hazardous <- factor(
  ifelse(as.character(model_data$Hazardous) %in% c("True", "TRUE"), "TRUE", "FALSE"),
  levels = c("FALSE", "TRUE")
)

```

## Train / Validation / Test Split — Random Sample Approach

```{r step2-split}
# Shared 60/20/20 stratified split for all models
set.seed(5703)
train_idx   <- createDataPartition(comets_model$Hazardous, p = 0.6, list = FALSE)
comet_train <- comets_model[train_idx, ]
temp        <- comets_model[-train_idx, ]

val_idx    <- createDataPartition(temp$Hazardous, p = 0.5, list = FALSE)
comet_val  <- temp[val_idx, ]
comet_test <- temp[-val_idx, ]

minority_n <- min(table(comet_train$Hazardous))

# Apply same split indices to hco_data for Step 2 models
train_hco_data      <- hco_data[train_idx, ]
validation_hco_data <- hco_data[val_idx, ]
test_hco_data       <- hco_data[-c(train_idx, val_idx), ]

str(hco_data)
str(comets_model)
table(comet_train$Hazardous)
```

## Baseline Logistic Regression

```{r step2-baseline}
baseline_model_hco_data <- glm(Hazardous ~ ., data = train_hco_data, family = binomial)
summary(baseline_model_hco_data)

kable(
  data.frame(
    Variable    = names(coef(baseline_model_hco_data)),
    Coefficient = round(as.numeric(coef(baseline_model_hco_data)), 6)
  ),
  caption   = "Baseline Logistic Regression: All Variables",
  col.names = c("Variable Name", "Estimated Coefficient"),
  align     = "l"
)
```

## Best Subset Selection

```{r step2-best-subset}
best_subset_hco_data <- regsubsets(Hazardous ~ .,
                                   data   = train_hco_data,
                                   nvmax  = 19,
                                   method = "exhaustive")

summary_best_subset_hco_data <- summary(best_subset_hco_data)
best_model_bic               <- which.min(summary_best_subset_hco_data$bic)
coefficients_best_model      <- coef(best_subset_hco_data, best_model_bic)

kable(
  data.frame(
    Vars  = 1:length(summary_best_subset_hco_data$adjr2),
    AdjR2 = round(summary_best_subset_hco_data$adjr2, 4),
    BIC   = round(summary_best_subset_hco_data$bic,   4),
    CP    = round(summary_best_subset_hco_data$cp,    4)
  ),
  caption   = "Best Subset Selection Performance Metrics",
  col.names = c("Number of Variables", "Adj R-Squared", "BIC", "Mallows' Cp"),
  align     = "c"
)

kable(
  data.frame(
    Variable    = names(coefficients_best_model),
    Coefficient = round(as.numeric(coefficients_best_model), 6)
  ),
  caption   = paste("Variables Selected by Best Subset Selection (Size =", best_model_bic, ")"),
  col.names = c("Variable Name", "Estimated Coefficient"),
  align     = "l"
)
```

```{r step2-best-subset-plots}
par(mfrow = c(1, 3))

plot(summary_best_subset_hco_data$adjr2,
     xlab = "Number of Variables", ylab = "Adjusted R²",
     type = "l", main = "Adjusted R²")
points(which.max(summary_best_subset_hco_data$adjr2),
       max(summary_best_subset_hco_data$adjr2), col = "red", cex = 2, pch = 20)

plot(summary_best_subset_hco_data$cp,
     xlab = "Number of Variables", ylab = "Cp",
     type = "l", main = "Cp")
points(which.min(summary_best_subset_hco_data$cp),
       min(summary_best_subset_hco_data$cp), col = "red", cex = 2, pch = 20)

plot(summary_best_subset_hco_data$bic,
     xlab = "Number of Variables", ylab = "BIC",
     type = "l", main = "BIC")
points(which.min(summary_best_subset_hco_data$bic),
       min(summary_best_subset_hco_data$bic), col = "red", cex = 2, pch = 20)

par(mfrow = c(1, 1))
```

## Stepwise Selection (Bidirectional)

```{r step2-stepwise}
train_hco_data$Hazardous      <- as.factor(train_hco_data$Hazardous)
validation_hco_data$Hazardous <- as.factor(validation_hco_data$Hazardous)
test_hco_data$Hazardous       <- as.factor(test_hco_data$Hazardous)

full_model_hco_data <- glm(Hazardous ~ ., data = train_hco_data, family = binomial)
stepwise_hco_data   <- stepAIC(full_model_hco_data, direction = "both", trace = FALSE)

summary(stepwise_hco_data)

kable(
  data.frame(
    Variable    = names(coef(stepwise_hco_data)),
    Coefficient = round(as.numeric(coef(stepwise_hco_data)), 6)
  ),
  caption   = "Variables Selected by Stepwise Selection (Bidirectional)",
  col.names = c("Variable Name", "Estimated Coefficient"),
  align     = "l"
)
```

## Ridge Regression

```{r step2-ridge}
train_hco_data$Hazardous      <- as.numeric(as.character(train_hco_data$Hazardous))
validation_hco_data$Hazardous <- as.numeric(as.character(validation_hco_data$Hazardous))
test_hco_data$Hazardous       <- as.numeric(as.character(test_hco_data$Hazardous))

predictors_train_hco_data      <- as.matrix(train_hco_data[, colnames(train_hco_data) != "Hazardous"])
target_train_hco_data          <- train_hco_data$Hazardous
predictors_validation_hco_data <- as.matrix(validation_hco_data[, names(validation_hco_data) != "Hazardous"])
target_validation_hco_data     <- validation_hco_data$Hazardous

set.seed(5703)
ridge_5_fold_hco_data <- cv.glmnet(predictors_train_hco_data,
                                   target_train_hco_data,
                                   alpha  = 0,
                                   family = "binomial",
                                   nfolds = 5)

ridge_lambda_min_hco_data  <- ridge_5_fold_hco_data$lambda.min
ridge_lambda_1se_hco_data  <- ridge_5_fold_hco_data$lambda.1se

predict_response_ridge_min <- predict(ridge_5_fold_hco_data,
                                      newx = predictors_validation_hco_data,
                                      s    = ridge_lambda_min_hco_data,
                                      type = "response")
predict_response_ridge_1se <- predict(ridge_5_fold_hco_data,
                                      newx = predictors_validation_hco_data,
                                      s    = ridge_lambda_1se_hco_data,
                                      type = "response")
predict_class_ridge_min    <- predict(ridge_5_fold_hco_data,
                                      newx = predictors_validation_hco_data,
                                      s    = ridge_lambda_min_hco_data,
                                      type = "class")

kable(
  data.frame(
    Metric = c("Lambda.min", "Lambda.1se", "Error (min)", "MSE (min)", "MSE (1se)"),
    Value  = c(
      round(ridge_lambda_min_hco_data, 6),
      round(ridge_lambda_1se_hco_data, 6),
      round(mean(predict_class_ridge_min != target_validation_hco_data), 6),
      round(mean((target_validation_hco_data - predict_response_ridge_min)^2), 6),
      round(mean((target_validation_hco_data - predict_response_ridge_1se)^2), 6)
    )
  ),
  align   = "c",
  caption = "Ridge Regression 5-fold Cross Validation"
) %>%
  kable_styling(position = "left", full_width = FALSE)

ridge_fit_hco_data <- glmnet(predictors_train_hco_data, target_train_hco_data,
                             alpha = 0, family = "binomial")
plot(ridge_fit_hco_data)
title("Ridge Regression: Coefficients vs. Log(Lambda)", line = 3)

roc_lambda_min_ridge <- roc(target_validation_hco_data,
                            predict_response_ridge_min[, 1], quiet = TRUE)
plot(roc_lambda_min_ridge, col = "blue", main = "Ridge ROC lambda.min")
legend("bottomright",
       legend = paste("Lambda.min (AUC =", round(auc(roc_lambda_min_ridge), 6), ")"))
```

## LASSO Regression

```{r step2-lasso}
set.seed(5703)
lasso_5_fold_hco_data <- cv.glmnet(predictors_train_hco_data,
                                   target_train_hco_data,
                                   alpha  = 1,
                                   family = "binomial",
                                   nfolds = 5)

lasso_lambda_min_hco_data  <- lasso_5_fold_hco_data$lambda.min
lasso_lambda_1se_hco_data  <- lasso_5_fold_hco_data$lambda.1se

predict_response_lasso_min <- predict(lasso_5_fold_hco_data,
                                      newx = predictors_validation_hco_data,
                                      s    = lasso_lambda_min_hco_data,
                                      type = "response")
predict_response_lasso_1se <- predict(lasso_5_fold_hco_data,
                                      newx = predictors_validation_hco_data,
                                      s    = lasso_lambda_1se_hco_data,
                                      type = "response")

lasso_fit_hco_data <- glmnet(predictors_train_hco_data, target_train_hco_data,
                             alpha = 1, family = "binomial")
plot(lasso_fit_hco_data)
title("LASSO Regression: Coefficients vs. Log(Lambda)", line = 3)

roc_lambda_min_lasso <- roc(target_validation_hco_data,
                            predict_response_lasso_min[, 1], quiet = TRUE)
plot(roc_lambda_min_lasso, col = "blue", main = "LASSO ROC lambda.min")
legend("bottomright",
       legend = paste("Lambda.min (AUC =", round(auc(roc_lambda_min_lasso), 6), ")"))
```

## Final Model Comparison — Step 2

```{r step2-final-comparison}
val_hazardous_numeric <- as.numeric(as.character(validation_hco_data$Hazardous))

best_vars <- names(coef(best_subset_hco_data, best_model_bic))
best_vars <- best_vars[best_vars != "(Intercept)"]
form_best <- as.formula(paste("Hazardous ~", paste(best_vars, collapse = " + ")))
fit_best  <- glm(form_best, data = train_hco_data, family = binomial)

pred_baseline <- predict(baseline_model_hco_data,
                         newdata = validation_hco_data, type = "response")
pred_stepwise <- predict(stepwise_hco_data,
                         newdata = validation_hco_data, type = "response")
pred_best     <- predict(fit_best,
                         newdata = validation_hco_data, type = "response")
pred_ridge    <- predict(ridge_5_fold_hco_data,
                         newx = predictors_validation_hco_data,
                         s    = ridge_lambda_min_hco_data, type = "response")[, 1]
pred_lasso    <- predict(lasso_5_fold_hco_data,
                         newx = predictors_validation_hco_data,
                         s    = lasso_lambda_min_hco_data, type = "response")[, 1]

roc_baseline <- roc(val_hazardous_numeric, pred_baseline, quiet = TRUE)
roc_stepwise <- roc(val_hazardous_numeric, pred_stepwise, quiet = TRUE)
roc_best     <- roc(val_hazardous_numeric, pred_best,     quiet = TRUE)
roc_ridge    <- roc(val_hazardous_numeric, pred_ridge,    quiet = TRUE)
roc_lasso    <- roc(val_hazardous_numeric, pred_lasso,    quiet = TRUE)

kable(
  data.frame(
    Method    = c("Baseline (Full Logistic)", "Stepwise (Bidirectional)",
                  "Best Subset", "Ridge (lambda.min)", "LASSO (lambda.min)"),
    Variables = c(length(coef(baseline_model_hco_data)) - 1,
                  length(coef(stepwise_hco_data)) - 1,
                  best_model_bic,
                  sum(coef(ridge_5_fold_hco_data, s = ridge_lambda_min_hco_data) != 0) - 1,
                  sum(coef(lasso_5_fold_hco_data, s = lasso_lambda_min_hco_data) != 0) - 1),
    AIC       = c(round(AIC(baseline_model_hco_data), 4),
                  round(AIC(stepwise_hco_data), 4),
                  round(AIC(fit_best), 4), NA, NA),
    AUC       = round(c(as.numeric(auc(roc_baseline)),
                        as.numeric(auc(roc_stepwise)),
                        as.numeric(auc(roc_best)),
                        as.numeric(auc(roc_ridge)),
                        as.numeric(auc(roc_lasso))), 4)
  ),
  caption   = "Final Model Comparison Across All Methods (Validation Set)",
  col.names = c("Method", "Variables Used", "AIC (Training)", "AUC (Validation)"),
  align     = "c"
) %>%
  kable_styling(position = "left", full_width = FALSE)

plot(roc_baseline, col = "black", lwd = 2,
     main = "ROC Curve Comparison: All Step 2 Models")
lines(roc_stepwise, col = "blue",   lwd = 2)
lines(roc_best,     col = "red",    lwd = 2)
lines(roc_ridge,    col = "green",  lwd = 2)
lines(roc_lasso,    col = "orange", lwd = 2)
legend("bottomright",
       legend = c(paste("Baseline    (AUC =", round(as.numeric(auc(roc_baseline)), 4), ")"),
                  paste("Stepwise    (AUC =", round(as.numeric(auc(roc_stepwise)), 4), ")"),
                  paste("Best Subset (AUC =", round(as.numeric(auc(roc_best)),     4), ")"),
                  paste("Ridge       (AUC =", round(as.numeric(auc(roc_ridge)),     4), ")"),
                  paste("LASSO       (AUC =", round(as.numeric(auc(roc_lasso)),     4), ")")),
       col = c("black", "blue", "red", "green", "orange"), lwd = 2)
```

---

# Ensemble Methods — Random Forest, Boosting, and Bagging

## Random Forest

### Model 1 — Baseline Random Forest

```{r rf-model-1}
set.seed(5703)

comet_rf_model <- randomForest(
  Hazardous  ~ .,
  data       = comet_train,
  ntree      = 500,
  mtry       = floor(sqrt(ncol(comet_train) - 1)),
  sampsize   = c("FALSE" = minority_n, "TRUE" = minority_n),
  importance = TRUE
)

comet_preds <- predict(comet_rf_model, newdata = comet_val)
confusionMatrix(comet_preds, comet_val$Hazardous, positive = "TRUE")
```

### mtry Tuning via OOB Error

```{r rf-tune}
set.seed(5703)

comet_tuned <- tuneRF(
  x          = comet_train[, -which(names(comet_train) == "Hazardous")],
  y          = comet_train$Hazardous,
  stepFactor = 1.5,
  improve    = 0.01,
  ntreeTry   = 300,
  trace      = TRUE,
  plot       = TRUE
)
```

### Model 2 — Tuned mtry with Class Weights

```{r rf-model-2}
set.seed(5703)

comet_rf_model_tune <- randomForest(
  Hazardous  ~ .,
  data       = comet_train,
  ntree      = 500,
  mtry       = 4,
  sampsize   = c("FALSE" = minority_n, "TRUE" = minority_n),
  classwt    = c("FALSE" = 1, "TRUE" = 10),
  importance = TRUE
)

comet_preds_1 <- predict(comet_rf_model_tune, newdata = comet_val)
confusionMatrix(comet_preds_1, comet_val$Hazardous, positive = "TRUE")
```

### Model 3 — Lowered Classification Threshold

```{r rf-model-3}
set.seed(5703)

comet_rf_model_tune_1 <- randomForest(
  Hazardous  ~ .,
  data       = comet_train,
  ntree      = 500,
  mtry       = 4,
  sampsize   = c("FALSE" = minority_n, "TRUE" = minority_n),
  importance = TRUE
)

pred_probs_1 <- predict(comet_rf_model_tune_1,
                        newdata = as.data.frame(comet_val), type = "prob")
pred_class_1 <- ifelse(pred_probs_1[, "TRUE"] >= 0.15, "TRUE", "FALSE")
pred_class_1 <- factor(pred_class_1, levels = c("FALSE", "TRUE"))

confusionMatrix(pred_class_1, comet_val$Hazardous, positive = "TRUE")
```

### ROC Curve — Optimal Threshold Selection

```{r rf-roc}
roc_obj <- roc(comet_val$Hazardous, pred_probs_1[, "TRUE"])
plot(roc_obj, print.thres = "best", print.auc = TRUE)
```

### Model 4 — ROC-Optimized Threshold (Final RF Model)

```{r rf-model-4}
set.seed(5703)

comet_rf_model_tune_2 <- randomForest(
  Hazardous  ~ .,
  data       = comet_train,
  ntree      = 500,
  mtry       = 4,
  sampsize   = c("FALSE" = minority_n, "TRUE" = minority_n),
  importance = TRUE
)

pred_probs_2 <- predict(comet_rf_model_tune_2,
                        newdata = as.data.frame(comet_val), type = "prob")
pred_class_2 <- ifelse(pred_probs_2[, "TRUE"] >= 0.528, "TRUE", "FALSE")
pred_class_2 <- factor(pred_class_2, levels = c("FALSE", "TRUE"))

confusionMatrix(pred_class_2, comet_val$Hazardous, positive = "TRUE")
```

### Final RF Test Set Evaluation

```{r rf-test}
set.seed(5703)

pred_probs_3 <- predict(comet_rf_model_tune_2,
                        newdata = as.data.frame(comet_test), type = "prob")
pred_class_3 <- ifelse(pred_probs_3[, "TRUE"] >= 0.528, "TRUE", "FALSE")
pred_class_3 <- factor(pred_class_3, levels = c("FALSE", "TRUE"))

confusionMatrix(pred_class_3, comet_test$Hazardous, positive = "TRUE")
```

---

## Gradient Boosting

### Data Recoding for Boosting Compatibility

```{r boost-recode}
comet_train_boost <- comet_train
comet_val_boost   <- comet_val
comet_test_boost  <- comet_test

comet_train_boost$Hazardous <- factor(
  ifelse(comet_train$Hazardous == "TRUE", "Yes", "No"), levels = c("No", "Yes"))
comet_val_boost$Hazardous   <- factor(
  ifelse(comet_val$Hazardous   == "TRUE", "Yes", "No"), levels = c("No", "Yes"))
comet_test_boost$Hazardous  <- factor(
  ifelse(comet_test$Hazardous  == "TRUE", "Yes", "No"), levels = c("No", "Yes"))
```

### Model Training — 5-Fold CV with Grid Search

```{r boost-train}
set.seed(5703)

gbm_grid <- expand.grid(
  interaction.depth = seq(1, 5, by = 1),
  n.trees           = seq(100, 600, by = 100),
  shrinkage         = c(0.001, 0.01, 0.1),
  n.minobsinnode    = c(10, 20)
)

n_models  <- nrow(gbm_grid)
setSeeds  <- vector(mode = "list", length = 6)
for (i in 1:5) setSeeds[[i]] <- sample.int(100000, n_models)
setSeeds[[6]] <- sample.int(100000, 1)

fitControl <- trainControl(
  method          = "cv",
  number          = 5,
  classProbs      = TRUE,
  summaryFunction = twoClassSummary,
  sampling        = "down",
  seeds           = setSeeds
)

cat("Starting cross-validation. This will take a moment...\n")
model_boost <- train(
  Hazardous    ~ .,
  data         = comet_train_boost,
  method       = "gbm",
  distribution = "bernoulli",
  trControl    = fitControl,
  tuneGrid     = gbm_grid,
  metric       = "ROC",
  verbose      = FALSE
)
```

### Optimal Tuning and Variable Importance

```{r boost-tune}
cat("OPTIMAL TUNING IN BOOSTING MODEL:\n")
print(model_boost$bestTune)
cat("Optimal number of trees from 5-fold CV:", model_boost$bestTune$n.trees, "\n")

plot(varImp(model_boost), top = 10, main = "Boosting: Variable Importance")
```

### Validation Set ROC — Threshold Selection

```{r boost-val-roc}
probs_boost_val <- predict(model_boost, newdata = comet_val_boost, type = "prob")[, "Yes"]
roc_boost_val   <- roc(response  = comet_val_boost$Hazardous,
                       predictor = probs_boost_val, quiet = TRUE)

plot(roc_boost_val, print.thres = "best", print.auc = TRUE,
     col = "blue", main = "ROC: Optimal Threshold of Validation Set")

optimal   <- coords(roc_boost_val, "best",
                    ret       = c("threshold", "specificity", "sensitivity"),
                    transpose = FALSE)
threshold <- optimal$threshold[1]
cat("Optimal Threshold Cutoff:", threshold, "\n")
```

### Validation Set Evaluation

```{r boost-val-eval}
preds_boost_val <- factor(ifelse(probs_boost_val >= threshold, "Yes", "No"),
                          levels = c("No", "Yes"))
cm_boost_val    <- confusionMatrix(preds_boost_val,
                                   comet_val_boost$Hazardous, positive = "Yes")
print(cm_boost_val)

cat("Accuracy: ",  round(cm_boost_val$overall["Accuracy"],  4), "\n")
cat("Precision: ", round(cm_boost_val$byClass["Precision"], 4), "\n")
cat("Recall: ",    round(cm_boost_val$byClass["Recall"],    4), "\n")
cat("F1 Score: ",  round(cm_boost_val$byClass["F1"],        4), "\n")
cat("ROC-AUC: ",   round(auc(roc_boost_val),                4), "\n")

plot(roc_boost_val, col = "forestgreen",
     main = "ROC Curve: Gradient Boosting (Validation)", lwd = 2)
plot(varImp(model_boost), top = 10,
     main = "Top 10 Predictors for Hazardous Asteroids")
```

### Test Set Evaluation

```{r boost-test-eval}
probs_boost_test <- predict(model_boost, newdata = comet_test_boost, type = "prob")[, "Yes"]
preds_boost_test <- factor(ifelse(probs_boost_test >= threshold, "Yes", "No"),
                           levels = c("No", "Yes"))

cm_boost_test <- confusionMatrix(preds_boost_test,
                                 comet_test_boost$Hazardous, positive = "Yes")
print(cm_boost_test)

roc_boost <- roc(response  = comet_test_boost$Hazardous,
                 predictor = probs_boost_test,
                 levels    = c("No", "Yes"),
                 direction = "<", quiet = TRUE)
ROC_AUC <- round(auc(roc_boost), 4)

cat("Accuracy: ",  round(cm_boost_test$overall["Accuracy"],  4), "\n")
cat("Precision: ", round(cm_boost_test$byClass["Precision"], 4), "\n")
cat("Recall: ",    round(cm_boost_test$byClass["Recall"],    4), "\n")
cat("F1 Score: ",  round(cm_boost_test$byClass["F1"],        4), "\n")
cat("ROC-AUC: ",   ROC_AUC,                                      "\n")

plot(roc_boost, col = "forestgreen",
     main = "ROC Curve: Gradient Boosting (Test)", lwd = 2)
plot(varImp(model_boost), top = 10,
     main = "Top 10 Predictors for Hazardous Asteroids")
```

---

## Bagging

### Model Fitting — 100, 300, 500 Trees

```{r bag-fit}
p <- ncol(comet_train) - 1

set.seed(5703)
bag_100 <- randomForest(Hazardous ~ ., data = comet_train, mtry = p, ntree = 100,
                        sampsize   = c("FALSE" = minority_n, "TRUE" = minority_n),
                        importance = TRUE)
set.seed(5703)
bag_300 <- randomForest(Hazardous ~ ., data = comet_train, mtry = p, ntree = 300,
                        sampsize   = c("FALSE" = minority_n, "TRUE" = minority_n),
                        importance = TRUE)
set.seed(5703)
bag_500 <- randomForest(Hazardous ~ ., data = comet_train, mtry = p, ntree = 500,
                        sampsize   = c("FALSE" = minority_n, "TRUE" = minority_n),
                        importance = TRUE)

bag_100
bag_300
bag_500
```

### OOB Error Plots

```{r bag-oob}
plot(bag_100, main = "Bagging OOB Error: 100 Trees")
plot(bag_300, main = "Bagging OOB Error: 300 Trees")
plot(bag_500, main = "Bagging OOB Error: 500 Trees")
```

### Validation Set Comparison

```{r bag-val}
get_classification_metrics <- function(model, newdata, truth_col = "Hazardous",
                                       positive_class = "TRUE", threshold = 0.5) {
  pred_prob  <- predict(model, newdata = newdata, type = "prob")[, positive_class]
  pred_class <- factor(ifelse(pred_prob >= threshold, "TRUE", "FALSE"),
                       levels = c("FALSE", "TRUE"))
  cm      <- confusionMatrix(pred_class, newdata[[truth_col]], positive = positive_class)
  roc_obj <- roc(response  = newdata[[truth_col]],
                 predictor = pred_prob,
                 levels    = c("FALSE", "TRUE"),
                 direction = "<")
  data.frame(
    Accuracy    = as.numeric(cm$overall["Accuracy"]),
    Precision   = as.numeric(cm$byClass["Pos Pred Value"]),
    Recall      = as.numeric(cm$byClass["Sensitivity"]),
    Specificity = as.numeric(cm$byClass["Specificity"]),
    F1          = as.numeric(cm$byClass["F1"]),
    AUC         = as.numeric(auc(roc_obj))
  )
}

bag_100_valid <- get_classification_metrics(bag_100, comet_val, threshold = 0.025)
bag_300_valid <- get_classification_metrics(bag_300, comet_val, threshold = 0.025)
bag_500_valid <- get_classification_metrics(bag_500, comet_val, threshold = 0.025)

validation_results <- rbind(
  cbind(Model = "Bagging_100", bag_100_valid),
  cbind(Model = "Bagging_300", bag_300_valid),
  cbind(Model = "Bagging_500", bag_500_valid)
)

kable(validation_results,
      caption = "Bagging Validation Set Results",
      align   = "c") %>%
  kable_styling(position = "left", full_width = FALSE)
```

### Final Bagging Model — Test Set Evaluation

```{r bag-test}
final_bag <- bag_300

bag_pred_class <- predict(final_bag, newdata = comet_test, type = "class")
bag_pred_prob  <- predict(final_bag, newdata = comet_test, type = "prob")[, "TRUE"]

bag_cm <- confusionMatrix(bag_pred_class, comet_test$Hazardous, positive = "TRUE")
bag_cm

bag_roc <- roc(response  = comet_test$Hazardous,
               predictor = bag_pred_prob,
               levels    = c("FALSE", "TRUE"),
               direction = "<")
bag_auc <- as.numeric(auc(bag_roc))
bag_auc

plot(bag_roc, col = "blue", main = "ROC Curve - Bagging Model")
varImpPlot(final_bag, main = "Variable Importance - Final Bagging Model")

bag_test_metrics <- data.frame(
  Model       = "Bagging",
  Accuracy    = as.numeric(bag_cm$overall["Accuracy"]),
  Precision   = as.numeric(bag_cm$byClass["Pos Pred Value"]),
  Recall      = as.numeric(bag_cm$byClass["Sensitivity"]),
  Specificity = as.numeric(bag_cm$byClass["Specificity"]),
  F1          = as.numeric(bag_cm$byClass["F1"]),
  AUC         = bag_auc
)

kable(bag_test_metrics,
      caption = "Bagging Final Test Set Metrics",
      align   = "c") %>%
  kable_styling(position = "left", full_width = FALSE)

bag_cm$table
```

---

# Model Comparison

## Confusion Matrices

```{r comparison-cm}
cat("=== Random Forest - Test Confusion Matrix ===\n")
print(table(pred_class_3, comet_test$Hazardous))

cat("\n=== Bagging - Test Confusion Matrix ===\n")
print(bag_cm$table)

cat("\n=== Boosting - Test Confusion Matrix ===\n")
print(cm_boost_test$table)
```

## ROC Curve Comparison

```{r comparison-roc}
roc_rf    <- roc(comet_test$Hazardous,       pred_probs_3[, "TRUE"],
                 levels = c("FALSE", "TRUE"), quiet = TRUE)
roc_bag   <- roc(comet_test$Hazardous,       bag_pred_prob,
                 levels = c("FALSE", "TRUE"), quiet = TRUE)
roc_boost <- roc(comet_test_boost$Hazardous, probs_boost_test,
                 levels = c("No", "Yes"),     quiet = TRUE)

plot(roc_rf,
     col  = "blue", lwd = 2,
     main = "ROC Curve Comparison: RF vs Bagging vs Boosting",
     legacy.axes = TRUE)
lines(roc_bag,   col = "darkgreen", lwd = 2)
lines(roc_boost, col = "red",       lwd = 2)
legend("bottomright",
       legend = c(paste("Random Forest (AUC =", round(auc(roc_rf),    4), ")"),
                  paste("Bagging       (AUC =", round(auc(roc_bag),   4), ")"),
                  paste("Boosting      (AUC =", round(auc(roc_boost), 4), ")")),
       col = c("blue", "darkgreen", "red"), lwd = 2)
```

## Metrics Comparison Table

```{r comparison-table}
comparison_table <- data.frame(
  Model       = c("Random Forest", "Bagging", "Boosting"),
  Accuracy    = c(
    round(confusionMatrix(pred_class_3, comet_test$Hazardous,
                          positive = "TRUE")$overall["Accuracy"], 4),
    round(bag_cm$overall["Accuracy"],        4),
    round(cm_boost_test$overall["Accuracy"], 4)
  ),
  Sensitivity = c(
    round(confusionMatrix(pred_class_3, comet_test$Hazardous,
                          positive = "TRUE")$byClass["Sensitivity"], 4),
    round(bag_cm$byClass["Sensitivity"],        4),
    round(cm_boost_test$byClass["Sensitivity"], 4)
  ),
  Specificity = c(
    round(confusionMatrix(pred_class_3, comet_test$Hazardous,
                          positive = "TRUE")$byClass["Specificity"], 4),
    round(bag_cm$byClass["Specificity"],        4),
    round(cm_boost_test$byClass["Specificity"], 4)
  ),
  Kappa = c(
    round(confusionMatrix(pred_class_3, comet_test$Hazardous,
                          positive = "TRUE")$overall["Kappa"], 4),
    round(bag_cm$overall["Kappa"],        4),
    round(cm_boost_test$overall["Kappa"], 4)
  ),
  F1 = c(
    round(confusionMatrix(pred_class_3, comet_test$Hazardous,
                          positive = "TRUE")$byClass["F1"], 4),
    round(bag_cm$byClass["F1"],        4),
    round(cm_boost_test$byClass["F1"], 4)
  ),
  AUC = c(
    round(as.numeric(auc(roc_rf)), 4),
    round(bag_auc,                 4),
    ROC_AUC
  )
)

kable(comparison_table,
      caption = "Final Model Comparison — Test Set Performance",
      align   = "c") %>%
  kable_styling(position = "center", full_width = FALSE)
```

## Variable Importance Comparison

```{r comparison-varimp}
rf_imp <- as.data.frame(importance(comet_rf_model_tune_2))
rf_imp$Variable <- rownames(rf_imp)
rf_imp <- rf_imp[, c("Variable", "MeanDecreaseGini")]
rf_imp$Model <- "Random Forest"

bag_imp <- as.data.frame(importance(final_bag))
bag_imp$Variable <- rownames(bag_imp)
bag_imp <- bag_imp[, c("Variable", "MeanDecreaseGini")]
bag_imp$Model <- "Bagging"

boost_imp <- as.data.frame(varImp(model_boost)$importance)
boost_imp$Variable <- rownames(boost_imp)
colnames(boost_imp)[colnames(boost_imp) == "Overall"] <- "MeanDecreaseGini"
boost_imp <- boost_imp[, c("Variable", "MeanDecreaseGini")]
boost_imp$Model <- "Boosting"
```

## Grouped Performance Bar Chart

```{r comparison-barchart}
comparison_long <- tidyr::pivot_longer(
  comparison_table[, c("Model", "Sensitivity", "Specificity", "AUC", "Kappa")],
  cols      = -Model,
  names_to  = "Metric",
  values_to = "Value"
)

ggplot(comparison_long, aes(x = Metric, y = Value, fill = Model)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = round(Value, 4)),
            position = position_dodge(width = 0.9),
            vjust    = -0.3,
            size     = 3) +
  scale_y_continuous(limits = c(0.95, 1.002), oob = scales::squish) +
  labs(title = "Model Performance Comparison — Test Set",
       x     = "Metric",
       y     = "Value") +
  theme_minimal() +
  scale_fill_manual(values = c("Random Forest" = "blue",
                               "Bagging"       = "darkgreen",
                               "Boosting"      = "red"))
```

```{r all-models-comparison-chart}
# Step 2 test set predictions
pred_baseline_test <- predict(baseline_model_hco_data, newdata = test_hco_data, type = "response")
pred_stepwise_test <- predict(stepwise_hco_data,        newdata = test_hco_data, type = "response")
pred_best_test     <- predict(fit_best,                 newdata = test_hco_data, type = "response")

predictors_test_hco_data <- as.matrix(test_hco_data[, names(test_hco_data) != "Hazardous"])
pred_ridge_test <- predict(ridge_5_fold_hco_data, newx = predictors_test_hco_data,
                           s = ridge_lambda_min_hco_data, type = "response")[, 1]
pred_lasso_test <- predict(lasso_5_fold_hco_data, newx = predictors_test_hco_data,
                           s = lasso_lambda_min_hco_data, type = "response")[, 1]

target_test_hco <- as.numeric(as.character(test_hco_data$Hazardous))

# ROC objects for Step 2 on test set
roc_baseline_test <- roc(target_test_hco, pred_baseline_test, quiet = TRUE)
roc_stepwise_test <- roc(target_test_hco, pred_stepwise_test, quiet = TRUE)
roc_best_test     <- roc(target_test_hco, pred_best_test,     quiet = TRUE)
roc_ridge_test    <- roc(target_test_hco, pred_ridge_test,    quiet = TRUE)
roc_lasso_test    <- roc(target_test_hco, pred_lasso_test,    quiet = TRUE)

# Helper to extract sensitivity/specificity at best threshold
get_best_coords <- function(roc_obj, metric) {
  round(as.numeric(coords(roc_obj, "best", ret = metric, transpose = FALSE)[1, ]), 4)
}

# Build unified comparison table
all_models_table <- data.frame(
  Model = c("Baseline Logistic", "Stepwise", "Best Subset",
            "Ridge", "LASSO",
            "Random Forest", "Bagging", "Boosting"),
  AUC = c(
    round(as.numeric(auc(roc_baseline_test)), 4),
    round(as.numeric(auc(roc_stepwise_test)), 4),
    round(as.numeric(auc(roc_best_test)),     4),
    round(as.numeric(auc(roc_ridge_test)),    4),
    round(as.numeric(auc(roc_lasso_test)),    4),
    round(as.numeric(auc(roc_rf)),            4),
    round(bag_auc,                            4),
    ROC_AUC
  ),
  Sensitivity = c(
    get_best_coords(roc_baseline_test, "sensitivity"),
    get_best_coords(roc_stepwise_test, "sensitivity"),
    get_best_coords(roc_best_test,     "sensitivity"),
    get_best_coords(roc_ridge_test,    "sensitivity"),
    get_best_coords(roc_lasso_test,    "sensitivity"),
    round(confusionMatrix(pred_class_3, comet_test$Hazardous,
                          positive = "TRUE")$byClass["Sensitivity"], 4),
    round(bag_cm$byClass["Sensitivity"],        4),
    round(cm_boost_test$byClass["Sensitivity"], 4)
  ),
  Specificity = c(
    get_best_coords(roc_baseline_test, "specificity"),
    get_best_coords(roc_stepwise_test, "specificity"),
    get_best_coords(roc_best_test,     "specificity"),
    get_best_coords(roc_ridge_test,    "specificity"),
    get_best_coords(roc_lasso_test,    "specificity"),
    round(confusionMatrix(pred_class_3, comet_test$Hazardous,
                          positive = "TRUE")$byClass["Specificity"], 4),
    round(bag_cm$byClass["Specificity"],        4),
    round(cm_boost_test$byClass["Specificity"], 4)
  )
)

# Pivot for plotting
all_models_long <- tidyr::pivot_longer(
  all_models_table,
  cols      = -Model,
  names_to  = "Metric",
  values_to = "Value"
)

all_models_long$Model <- factor(all_models_long$Model,
  levels = c("Baseline Logistic", "Stepwise", "Best Subset",
             "Ridge", "LASSO", "Random Forest", "Bagging", "Boosting"))

# Facet by metric to fix squishing
ggplot(all_models_long, aes(x = Model, y = Value, fill = Model)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = round(Value, 3)),
            vjust = -0.3, size = 2.5) +
  facet_wrap(~ Metric, ncol = 3) +
  scale_y_continuous(limits = c(0.85, 1.02), oob = scales::squish) +
  labs(title = "Model Performance Comparison — All Models (Test Set)",
       x     = NULL,
       y     = "Value") +
  theme_minimal() +
  theme(axis.text.x  = element_text(angle = 45, hjust = 1),
        legend.position = "none") +
  scale_fill_manual(values = c(
    "Baseline Logistic" = "#E41A1C",
    "Stepwise"          = "#FF7F00",
    "Best Subset"       = "#FFFF33",
    "Ridge"             = "#4DAF4A",
    "LASSO"             = "#377EB8",
    "Random Forest"     = "blue",
    "Bagging"           = "darkgreen",
    "Boosting"          = "red"
  ))
```

